---
title: "CSSS508, Lecture 7"
subtitle: "Functions"
author: "Michael Pearce<br>(based on slides from Chuck Lanfear)"
date: "November 10, 2022"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["center","top"]
---
class:inverse
```{r setup, include=FALSE, purl=FALSE}
options(htmltools.dir.version = FALSE)
options(width=70)
knitr::opts_chunk$set(comment = "##")
```

# Reminders

+ Homework 6 is due at *midnight* tonight!

+ Lab/Office Hours for next week are *modified:*

  + Office Hours: Monday (11/14) 8:30-10:30am; Thursday (11/17) 10am-12pm
  + Lab (11/15): Replaced with additional office hours (above!)
  
+ Homework 7 will be due on on 11/17 at *midnight*

  + You will have time in class today to work on this homework!
  + *No assignment due Thanksgiving!*

---
class:inverse
# Topics

Last time, we learned about,

1. Why we use loops
1. `for()` loops
1. `while()` loops

--

Today, we will cover,

1. Aside: Visualizing the Goal
1. Building blocks of functions
1. Simple functions
1. Using functions with `apply()`

---
class:inverse

# 1. Visualizing the Goal

---
## Visualizing the Goal

Before you can write effective code, you need to know *exactly* what you want:

+ **Goal:** Do I want a single value? vector? one observation per person? per year?

--

+ **Current State:** What do I currently have? matrix, vector? long or wide format?

--

+ **Translate:** How can I take what I have and turn it into my goal?

  + Sketch out the steps!
  + Break it down into little operations
  
--

**As we become more advanced coders, this concept is key!!**

**Remember:** *When you're stuck, try searching your problem on Google!!*


---
class: inverse
# 2. Building blocks of functions

---
## Why Functions?

R (as well as mathematics in general) is full of functions! 

--

We use functions to:

  + Compute summary statistics (`mean()`, `sd()`, `min()`)
  + Fit models to data (`lm(Fertility~Agriculture,data=swiss)`)
  + Load data (`read_csv()`)
  + Create ggplots (`ggplot()`)
  + And so much more!!
  
---
## Examples of Existing Functions

* `mean()`:
    + Input: a vector
    + Output: a single number

--

* `dplyr::filter()`:
    + Input: a data frame, logical conditions
    + Output: a data frame with rows removed using those conditions

--

* `readr::read_csv()`:
    + Input: a file path, optionally variable names or types
    + Output: a data frame containing info read in from file
  
--

Each function requires **inputs**, and returns **outputs**

---
## Why Write Your Own Functions?

Functions encapsulate actions you might perform often, such as:

* Given a vector, compute some special summary stats
* Given a vector and definition of "invalid" values, replace with `NA`
* Defining a new logical operator

--

Advanced function applications (not covered in this class):

* Parallel processing
* Generating *other* functions
* Making custom packages containing your functions

---
## Anatomy of a Function


```{r, eval=FALSE}
NAME <- function(ARGUMENT1, ARGUMENT2=DEFAULT){
  BODY
  return(OUTPUT)
}
```


* **Name**: What you call the function so you can use it later

--

* **Arguments** (aka inputs, parameters): things the user passes to the function that affect how it works
    + e.g. `ARGUMENT1`, `ARGUMENT2`
    + `ARGUMENT2=DEFAULT` is example of setting a default value
    + In this example, `ARGUMENT1`, `ARGUMENT2` values won't exist outside of the function
--

* **Body**: The actual operations inside the function.
--

* **Output**: The object inside `return()`. Could be anything (or nothing!)
    + If unspecified, will be the last thing calculated
    

---
class: inverse
# 3. Simple functions

---
## Example 1: Doubling A Number

```{r}
double_x <- function(x){
  double_x <- x * 2
  return(double_x)
}
```

--

Let's run it!

```{r}
double_x(5)
double_x(NA)
double_x(1:2)
```
---
## Example 2: Extract First/Last

```{r}
first_and_last <- function(x) {
    first <- x[1]
    last  <- x[length(x)]
    return(c("first" = first, "last" = last))
}
```

--

Test it out:

```{r}
first_and_last(c(4, 3, 1, 8))
```


---
## Example 2: Testing `first_and_last`

What if I give `first_and_last()` a vector of length 1?

```{r}
first_and_last(7)
```

--

Of length 0?

```{r}
first_and_last(numeric(0))
```

--

Maybe we want it to be a little smarter.

---
## Example 3: Checking Inputs

Let's make sure we get an error message when the vector is too small:

```{r}
smarter_first_and_last <- function(x) {
    if(length(x) < 2){
      stop("Input is not long enough!")
    } else{
      first <- x[1]
      last  <- x[length(x)]
      return(c("first" = first, "last" = last))    
    }
}
```

.footnote[`stop()` ceases running the function and prints the text inside as an error message.]
---
## Example 3: Testing Smarter Function

```{r, error=TRUE}
smarter_first_and_last(NA)
smarter_first_and_last(c(4, 3, 1, 8))
```

---
## Cracking Open Functions

If you type a function name without any parentheses or arguments, you can see its contents:

```{r}
smarter_first_and_last
```

---
class:inverse

## Activity: Writing A Function

In Olympic diving, a panel of 7 judges provide scores. After removing the worst and best scores, the mean of the remaining scores is given to the diver. We'll write code to calculate this score!

1. Suppose I give you an ordered vector (lowest to greatest) of length 7. How can you keep elements 2 through 6 (i.e., remove the first and last values).

2. Write a function to calculate a diver's score:
  + Input: Ordered vector of length 7
  + Output: Mean score after removing the first and last scores.
  + Checks: Check that the vector has length 7.
  
3. Calculate the diver's score given `x <- c(1,3:7,10)`

---
## Activity: My Solution

1. Extract elements 2 through 6:

  + **Answer:** Given vector `x`, use `x[2:6]`
  
--

2. Function

```{r}
divers_score <- function(x){
  if(length(x) != 7){
    stop("x is not of length 7!")
  } else{
    x_nofirst_nolast <- x[2:6]
    return(mean(x_nofirst_nolast))
  }
}
```

--

3. Calculate the diver's score given `x <- c(1,3:7,10)`

```{r}
divers_score(x = c(1,3:7,10))
```

---
class: inverse
# 4. Using functions with `apply()`

---
## Applying Functions Multiple Times?

Last week, we saw an example where we wanted to take the mean of each column in the `swiss` data:

```{r}
for(col_index in seq_along(swiss)){
  mean_swiss_col <- mean(swiss[,col_index])
  names_swiss_col <- names(swiss)[col_index]
  print(c(names_swiss_col,round(mean_swiss_col,3)))
}
```

*Isn't this kind of complex?!*

---
## `apply()`, don't loop!

Writing loops can be challenging and prone to bugs!!

--

The `apply()` can solve this issue:

+ **apply** a function to values in each row or column of a matrix
+ Doesn't require preallocation
+ Can take built-in functions or user-created functions.

---
## Structure of `apply()`

`apply()` takes 3 arguments:

1. Data (a matrix or data frame)
2. Margin (1 applies function to each *row*, 2 applies to each *column*)
3. Function


```{r eval=FALSE}
apply(DATA, MARGIN, FUNCTION)
```

--

For example,

```{r}
apply(swiss, 2, mean)
```

---
## Example 1


```{r}
row_max <- apply(swiss,1,max) #maximum in each row
head(row_max,20)
```

---
## Example 2

```{r}
apply(swiss,2,summary) # summary of each column
```

**Note:* Matrix output!

---
## Example 3: User-Created Function

```{r}
scores <- matrix(1:21,nrow=3)
print(scores)
apply(scores,1,divers_score)
```

---
class:inverse
## Activity

*These are homework questions!!*

1. What's similar/different between a `for` loop and `apply`?

1. Write an `apply()` function to take the median value of each column in the `cars` dataset

1. Using `ggplot`, make a scatterplot of the `speed` and `dist` variables in `cars`. Then, add an appropriate horizontal and vertical line symbolizing the median value of each variable.

*Hint:* Using the layers `geom_vline(xintercept = )` and `geom_hline(yintercept = )`

---
## My Answers

1. What's similar/different between a `for` loop and `apply`?

  + **Possible Answer:** `for` loops and `apply` can usually accomplish the same tasks: Each uses code iteratively. 
  + `for` loops are more flexible, in that you can provide different data using indices to the loop; they are harder to write. 
  + `apply` loops are simpler and don't require preallocation; they require your data be a matrix or dataframe.

--

1. Write an `apply()` function to take the median value of each column in the `cars` dataset

```{r}
median_cars <- apply(cars,2,median)
median_cars
```

---
## My Answers

3. Make a ggplot


```{r fig.height=4}
library(ggplot2)
ggplot(cars,aes(speed,dist))+geom_point()+
  geom_vline(xintercept = median_cars[1])+
  geom_hline(yintercept = median_cars[2])
```


---
class: inverse
# Homework

Time to work on Homework 7!

